Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-23 manuscript.Rmd
Discussion I

A recurring focus in the sociology of culture is the question of whether people, as they move through their lives and transition between social roles, change their personal culture –- the declarative and non-declarative attitudes, beliefs, values, practices, and dispositions that exist at the individual level [@lizardo2017; @kiley2020; @lersch2023].

@@

These seemingly inconsistent findings are partly due to the fact that researchers have taken a “tournament of models” approach in adjudicating the different perspectives [@lersch2023: p. 228]. Accordingly, researchers have formalized models of responses to cultural items over time as demonstrating either “change” or “no change” to then compared which model fits observed data better, thus functionally asking whether we ever observe within-person change in personal culture.

@@

This paper set out to clarify previous findings about the prevalence of durable change in personal culture -- the attitudes, beliefs, values, preferences, and behaviors that manifest at the individual level. Relying on an approach that quantified the amount of systematic variance attributable to either intrapersonal change or interpersonal differences at baseline, we found a consistent pattern across datasets previously used in this debate. We see evidence of intrapersonal change over time on almost all questions, with some measures of personal culture displaying quite high levels of intrapersonal change relative to the amount of variance between people. On some questions, such as the factors that affect one’s life satisfaction or attitudes about government benefits, it is highly plausible that differences in peoples’ experiences as adults are the primary explanation for differences in personal culture.

However, the amount of intrapersonal change is often small relative to interpersonal differences (less than 10 percent on average across all surveys). On many questions, the amount of systematic variance attributable to intrapersonal change is essentially zero. These include questions about civil liberties, abortion, generalized trust, or sense of civic duty. On these questions and the large number of others with a low proportion of systematic variance attributable to intrapersonal change, there does not appear to be sufficient cultural change during adulthood for experiences and transitions between social locations to be the primary explanation for why adults end up with different personal cultures. For these items, the explanation for personal cultural differences seems to reside principally in early-life experiences.

More important than adjudicating past debates, however, we show that the approach developed in this paper can adjudicate novel theoretical questions about the conditions in which change or stability is more prevalent and more important for explaining cultural differences within a population. As an example, we showed that the role of intrapersonal change in explaining cultural differences varies significantly by question and by education. Intrapersonal change plays almost no role in explaining general sentiments about the government's role in the economy, while it plays a much larger role in explaining ideological and partisan identification. At the same time, across issues, intrapersonal change is much more important for explaining cultural variation among people without a college degree than for people with a college degree. While this pattern might be confounded by age and other factors that affect college completion, it potentially suggests that college-going locks in personal culture in a way that other experiences do not, making adult experiences less influential for political dispositions for college graduates compared to the rest of the population.

>>>>

A recurring focus in the sociology of culture is the question of whether people, as they move through their lives, change their personal culture –- the declarative and non-declarative attitudes, beliefs, values, practices, and dispositions [@lizardo2017; @kiley2020; @lersch2023].

@@

These seemingly inconsistent findings are partly due to the fact that researchers have taken a “tournament of models” approach in adjudicating the different perspectives [@lersch2023: p. 228]. Accordingly, researchers have formalized models of responses to cultural items over time as demonstrating either “change” or “no change” to then compare which model fits observed data better, thus functionally asking whether we ever observe within-person change in personal culture.

@@

This paper set out intervene in the debate on whether people change their personal culture -- their attitudes, beliefs, values, practices, and dispositions -- as they move through life. Instead of falling back on a "tournament of models" approach [@lersch2023: p. 228] to call victory for one or the other possible answer to this question, this paper seeks to refocus the debate to the theoretically more productive question of _how_ initial differences between and durable changes within people collectively explain the observed variation in personal culture. To do so, it proposes a new measure that quantifies the amount of systematic variance that is attributable to either interpersonal differences at baseline or intrapersonal change.

Applying this measure to all datasets previously discussed in this debate revealed a consistent pattern. Nearly all questions on personal culture show evidence of intrapersonal change over time, with some items showing noteably high intrapersonal change compared to interpersonal variance. For questions about factors that affect one’s life satisfaction or views on government benefits, it seems plausible that differences in adult experiences predominantly account for the observed differences between people.

Intrapersonal change is often substantially less pronounced than interpersonal differences, accounting for less than 10 percent on average across all surveys. On numerous questions, like those on civil liberties, abortion, generalized trust, and civic duty, the systematic variance attributable to intrapersonal change is essentially zero. For these and other questions with minimal systematic variance attributable from intrapersonal change, it seems there is not enough cultural change during adulthood to attribute differences in personal culture mainly to experiences and social transitions. For these topics, the primary source of persona cultural differences appeasr to stem from early-life experiences.

More important than adjudicating past debates, however, the approach presented in this paper offers insights into new theoretical inquiries regarding when change or stability in personal culture predominates and its significance in explaining cultural variances within a population.

For instance, we detailed how the contribution that intrapersonal change can make in understanding cultural differences varies substantially by question and by individuals' education. It plays almost no part in explaining sentiments regarding the government's economic role but is more defining for ideological and partisan identities. Intriguingly, intrapersonal change is more important in explaining cultural variation among people without a college compared to college graduates in the U.S.. While age and other factors related to college completion might confound this pattern, it suggests that college may solidify personal culture to an extent that renders the effect of other adult experiences on political dispositions for graduates less influential relative to individuals without such a college experience.
